Bulk Semantic Annotation with a Partially Known Knowledge Base

### 2024-11-28 doc/cicm25-sysdesc/sysdesc.tex
some work on sysdesc

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=10cm]{../example2.pdf}}
  \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
\end{figure}

@@

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm]{../search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

The \snify system (see \cite{stextools:git}) is a simple command-line tool that creates a
database of symbol-verbalization pairs, analyzes \sTeX source files and then steps through
all word occurrences in the document that share the stem with a verbalization in the
database. For each such word, the user is presented with an annotation choice and
interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
shows a typical situation.\ednote{MK: I like the example less and less! We need an example
  that has at least one homostem (another verbalization that has the same stem; introduce
  the concept her and use it throughout) so that we can show the choice mechanism, and
  where the verbalization is different from the symbol name. Having a two-word compound is
  good though.}

\begin{figure}[ht]
  \fbox{\includegraphics[width=12cm]{../img/srify}}
  \caption{\snify in Action: Annotating the term ``state space''.}\label{fig:snify}
\end{figure}

Here the annotator can choose an annotation by typing the corresponding choice
number. Here, the module \lstinline|state-space| is already imported (see the little green
check-mark); otherwise \snify would offer the annotator the choice to import it -- if we are
in a module context -- or to add a \lstinline|\usemodule| directive and where (in the
local environment or top-level). Alternatively, typing \lstinline|s| skips this word,
\lstinline|S| skips and adds a comment to always skip this word, and -- if \snify first
suggests the term `state` \lstinline|n| extends it to the compound ``state space'' by
including the next word.

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
lecture notes; then the session consists of all included files. Other productivity
features include a focus mode that can be used to propagate a particular choice to the
rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
annotation This reduces the cognitive effort from choosing a symbol from a list to
deciding whether a particular choice is applicable in the current document context. Of
course the include management still needs to be done locally.

@@

The current interface is not well-suited for on-the-fly annotation while editing while
authoring. For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.

@@

In this system description we have presented the \snify tool, a command-line tool
\ednote{continue} \snify is open source\ednote{MK@FS: do we have an explicit license yet in
  the repository.}; the source and documentation are available from \cite{stextools:git}.

>>>>

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
  \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
\end{figure}

@@

\begin{figure}[ht]\centering
  \fbox{\includegraphics[width=12cm, trim={1cm 0 0 0}, clip]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

The \snify system (see \cite{stextools:git}) is a simple command-line tool.
It creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
and then steps through all word occurrences in the document that match a
verbalization in the database.
Matching is based on word stems using off-the-shelf stemmers, which allows us to
e.g.\ match ``utilities'' with ``utility''.
For each matched word, the user is presented with an annotation choice and
interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
shows a typical situation.

\begin{figure}[ht]
  \setlength{\fboxsep}{0pt}
  \fbox{\includegraphics[width=12cm,trim={0 3cm 0 0},clip]{../img/snify}}
  \caption{\snify in Action: Annotating the word ``terminal''.}\label{fig:snify}
\end{figure}

The annotator can choose an annotation by typing the corresponding choice number,
in this case \lstinline|0|.
The little green check-mark indicates that the relevant module is already imported.
Otherwise \snify would offer the annotator the choice to import it -- if we are
in a module context -- or to add a \lstinline|\usemodule| directive, either in the
local environment or at the top-level.

To skip the word, the user can type \lstinline|s|,
\lstinline|s!| skips it until the end of the file, and \lstinline|S|
will skip it in future runs as well by appending a comment to the file.
There are numerous other commands, e.g.\ to adjust the selection,
search for alternative annotation targets,
or view the file introducing one of the choices.

\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
together form a \textbf{session}, which can be interrupted and resumed without having to
re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
lecture notes; then the session consists of all included files. Other productivity
features include a focus mode that can be used to only annotate a particular word in the
rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
annotation. This reduces the cognitive overhead from switching between different 
words and symbol lists.
% Of course the include management still needs to be done locally.

@@

The current interface is not well-suited for on-the-fly annotation while authoring.
For that, the underlying information (the symbol/verbalization pairs harvested
from the domain model) can be integrated into any IDE. In fact we plan to do this for the
next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.

@@

In this system description we have presented \snify, a command-line tool
\ednote{continue} \snify is open source;
the source and documentation are available from \cite{stextools:git}.
